Likelihoods for a general class of ARGs under the SMC

### 2025-02-21 paper.tex
Full pass through intro

In this paper we provide a new formulation of ARG likelihood which 
has much more lenient input requirements than KYF, and gracefully handles
incomplete and imprecise ARGs.
% In this paper we provide a necessary first step towards this goal
% of statistically rigorous large-scale ARG inference, by defining a function
% to compute the likelihood of a general class of ARG under the SMC.
To do this, we give a backwards-in-time description of the SMC
which naturally leads to a straightforward method for computing the likelihood
of any ARG, and describe an efficient algorithm to do so.
We also illustrate the utility of this approach via some simple examples.

@@

Our final example explores decomposition of the likelihood
into separate components for different segments of time.
By explicitly associating a single likelihood with each edge,
we can compute the likelihood for any slice of the ARG taken in
time and/or along the genome.
In the case of a fully precise ARG, in which each recombination
event has an exact time,
this is true as long as the slicing operation does not erase the
information on whether an edge is associated with the left or
right hand segment generated by a recombination event. In the absence
of the exact time to recombination, the integral in
eq.~\eqref{eq:depth} means that the likelihood
for the entire ARG does not factorize into the likelihoods of these
individual slices. More precisely, the likelihood associated with any
edge that is initiated by a recombination breakpoint cannot be factorized
into the product of two integrals of the same form.
However, as long as the individual slices of the ARG contain sufficient
information we can approximate the true integral as the product of the
integral of the slices. Fig.~\ref{fig:3-arg-slices} demonstrates this idea
on a toy example. We simulated an ARG under human-like parameters with
two effective population size changes, and subsequently computed the posterior
distribution on the three effective population sizes
using a uniform prior on each slice independently.
The result is reasonably accurate, despite only using 10kb of sequence.

@@

% % % % % % % % % % % % % % % % % %
\subsection*{Algorithm} \label{par:algo}
In this section we describe how to efficiently compute the likelihood under the SMC
from an ARG stored using the \texttt{tskit} tree sequence encoding. 
The tree sequence format encodes relationships using \emph{edges},
along with additional information that
allows us to sequentially recover (and reason about 
their differences) 
efficiently~\citep{kelleher_efficient_2016, ralph_efficiently_2020}.
The order of the edge deletions and insertions required to move from one local tree $T_1$ to
the next, $T_2$ imposes the strict total ordering on edges
(see definition \eqref{def:coal})
that is needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
Indeed, following the removal of edges that do not persist in $T_2$,
we simply declare that each subsequently added edge
can only coalesce with the already-visited edges 
that make up the (partially) reconstructed local tree $T_2$.

@@

This allows us to compute a likelihood in a single pass over all edges.
More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
the number of extant lineages in each internode interval (see Fig.~\ref{fig:algo}).
When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along 
the internode intervals
spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
For each subsequently inserted edge, we first compute the likelihood contribution
of that edge by computing the integral
in equation \eqref{eq:full-lik} using $I$, and then update the counts in $I$.
We further keep track of
the last parent a child node was connected to to detect recombination events.

>>>>

In this paper we provide a new formulation of ARG likelihood which 
has much more lenient input requirements than KYF, and gracefully handles
incomplete and imprecise ARGs.
% In this paper we provide a necessary first step towards this goal
% of statistically rigorous large-scale ARG inference, by defining a function
% to compute the likelihood of a general class of ARG under the SMC.
To do this, we give a backwards-in-time description of the SMC
which naturally leads to a straightforward method for computing the likelihood
of any ARG, and describe an efficient algorithm to do so.
We show that explicitly identifying recombination events via nodes 
in the graph is not necessary as long as the ``locally
unary''~\cite{wong_general_2023,fritze2024forest} sections of common ancestor
nodes are retained, from the perspective of computing the likelihood.
We also show via a empirical examples that the likelihood is remarkably
robust to the presence of polytomies, and illustrate how it can be
decomposed into separately evaluated time slices.
Finally, we discuss how these results may be applied and extended.

@@

Our final example explores decomposition of the likelihood
into separate components for different segments of time.
By explicitly associating a single likelihood with each edge,
we can compute the likelihood for any slice of the ARG taken in
time and/or along the genome.
In the case of a fully precise ARG, in which each recombination
event has an exact time,
this is true as long as the slicing operation does not erase the
information on whether an edge is associated with the left or
right hand segment generated by a recombination event. In the absence
of the exact time to recombination, the integral in
Eq.~\eqref{eq:depth} means that the likelihood
for the entire ARG does not factorize into the likelihoods of these
individual slices. More precisely, the likelihood associated with any
edge that is initiated by a recombination breakpoint cannot be factorized
into the product of two integrals of the same form.
However, as long as the individual slices of the ARG contain sufficient
information we can approximate the true integral as the product of the
integral of the slices. Fig.~\ref{fig:3-arg-slices} demonstrates this idea
on a toy example. We simulated an ARG under human-like parameters with
two effective population size changes, and subsequently computed the posterior
distribution on the three effective population sizes
using a uniform prior on each slice independently.
The result is reasonably accurate, despite only using 10kb of sequence.

@@

% % % % % % % % % % % % % % % % % %
\subsection*{Algorithm} \label{par:algo}
In this section we describe how to efficiently compute the likelihood under the SMC
from an ARG stored using the \texttt{tskit} tree sequence encoding. 
The tree sequence format encodes relationships using \emph{edges},
along with additional information that
allows us to sequentially recover (and reason about 
their differences) 
efficiently~\citep{kelleher_efficient_2016, ralph_efficiently_2020}.
The order of the edge deletions and insertions required to move from one local tree $T_1$ to
the next, $T_2$ imposes the strict total ordering on edges
(see definition~\eqref{def:coal})
that is needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
Indeed, following the removal of edges that do not persist in $T_2$,
we simply declare that each subsequently added edge
can only coalesce with the already-visited edges 
that make up the (partially) reconstructed local tree $T_2$.

@@

This allows us to compute a likelihood in a single pass over all edges.
More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
the number of extant lineages in each internode interval (see Fig.~\ref{fig:algo}).
When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along 
the internode intervals
spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
For each subsequently inserted edge, we first compute the likelihood contribution
of that edge by computing the integral
in Eq.~\eqref{eq:full-lik} using $I$, and then update the counts in $I$.
We further keep track of
the last parent a child node was connected to to detect recombination events.
